---
title: "CS01: Biomarkers of Recent Use"
author: "Derek, Eddie, Rainer"
output: 
  html_document:
    toc: true
    toc_float: true
---

## Introduction

```{r setup, include=FALSE}
# control global Rmd chunk settings
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```

### Load packages

```{r load-packages, message=FALSE}
library(tidymodels)
library(tidyverse)
library(dplyr)
library(ggplot2)
library(janitor)
```

## Question

-   Which compound, in which matrix, and at what cutoff is the best biomarker of recent use?
-   Is the placebo associated with the release of chemical components within participants' bodies, leading them to perceive a sensation of being "high"?

## The Data

### Data Import

```{r}
Blood = read.csv("data/Blood.csv")
Breath = read_csv("data/Breath.csv")
OF = read.csv("data/OF.csv")

```

### Data Wrangling

We recoded and releveled variables (`Treatment` and `Group`), cleans column names, and renames specific columns (`x11_oh_thc` to `thcoh`, `thc_v` to `thcv`, `thccooh_gluc` to `thc_cooh_gluc`, and `thccooh` to `thc_cooh`) in all 3 tables. Using `janitor` package to organized column names.

```{r}
OF <- OF |>
  mutate(Treatment = fct_recode(Treatment, 
                                "5.9% THC (low dose)" = "5.90%",
                                "13.4% THC (high dose)" = "13.40%"),
         Treatment = fct_relevel(Treatment, "Placebo", "5.9% THC (low dose)"),
         Group = fct_recode(Group, 
                            "Occasional user" = "Not experienced user",
                            "Frequent user" = "Experienced user" )) |>  
  janitor::clean_names() |>
  rename(thcoh = x11_oh_thc,
         thcv = thc_v)
OF

Blood <- Blood |> 
  mutate(Treatment = fct_recode(Treatment, 
                                "5.9% THC (low dose)" = "5.90%",
                                "13.4% THC (high dose)" = "13.40%"),
         Treatment = fct_relevel(Treatment, "Placebo", "5.9% THC (low dose)")) |> 
  janitor::clean_names() |>
  rename(thcoh = x11_oh_thc,
         thccooh = thc_cooh,
         thccooh_gluc = thc_cooh_gluc,
         thcv = thc_v)
Blood

Breath <- Breath |> 
  mutate(Treatment = fct_recode(Treatment, 
                                "5.9% THC (low dose)" = "5.90%",
                                "13.4% THC (high dose)" = "13.40%"),
         Treatment = fct_relevel(Treatment, "Placebo", "5.9% THC (low dose)"),
         Group = fct_recode(Group, 
                            "Occasional user" = "Not experienced user",
                            "Frequent user" = "Experienced user" )) |> 
  janitor::clean_names() |> 
  rename(thc = thc_pg_pad)
Breath
```

Created 3 tables based on specific minutes and labeled accordingly, covering pre-smoking and subsequent post-smoking time periods for blood, breath, and oral fluid data.

```{r}
timepoints_Blood <- tibble(
  start = c(-400, 0, 30, 70, 100, 180, 210, 240, 270, 300),
  stop = c(
    0,
    30,
    70,
    100,
    180,
    210,
    240,
    270,
    300,
    max(Blood$time_from_start, na.rm = TRUE)
  ),
  timepoint = c(
    "pre-smoking",
    "0-30 min",
    "31-70 min",
    "71-100 min",
    "101-180 min",
    "181-210 min",
    "211-240 min",
    "241-270 min",
    "271-300 min",
    "301+ min"
  )
)

timepoints_Breath <- tibble(
  start = c(-400, 0, 40, 90, 180, 210, 240, 270),
  stop = c(
    0,
    40,
    90,
    180,
    210,
    240,
    270,
    max(Breath$time_from_start, na.rm = TRUE)
  ),
  timepoint = c(
    "pre-smoking",
    "0-40 min",
    "41-90 min",
    "91-180 min",
    "181-210 min",
    "211-240 min",
    "241-270 min",
    "271+ min"
  )
)

timepoints_OF <- tibble(
  start = c(-400, 0, 30, 90, 180, 210, 240, 270),
  stop = c(0, 30, 90, 180, 210, 240, 270,
           max(OF$time_from_start, na.rm = TRUE)),
  timepoint = c(
    "pre-smoking",
    "0-30 min",
    "31-90 min",
    "91-180 min",
    "181-210 min",
    "211-240 min",
    "241-270 min",
    "271+ min"
  )
)

assign_timepoint <- function(x, timepoints) {
  if (!is.na(x)) {
    timepoints$timepoint[x > timepoints$start & x <= timepoints$stop]
  } else{
    NA
  }
}
```

We created a new column, **`timepoint_use`**, in each table by mapping the **`time_from_start`** values to specific timepoints defined in separate reference data frames (**`timepoints_WB`**, **`timepoints_OF`**, **`timepoints_BR`**). Finally, releveled the **`timepoint_use`** factor variable to align with the order specified in the reference data frames. This ensures consistent and meaningful timepoint labels for subsequent analyses or visualizations in the study.

```{r}
Blood <- Blood |> 
  mutate(
    timepoint_use = map_chr(time_from_start,
                            assign_timepoint,
                            timepoints = timepoints_WB),
    timepoint_use = fct_relevel(timepoint_use, timepoints_WB$timepoint)
  )

OF <- OF |> 
  mutate(
    timepoint_use = map_chr(time_from_start,
                            assign_timepoint,
                            timepoints = timepoints_OF),
    timepoint_use = fct_relevel(timepoint_use, timepoints_OF$timepoint)
  )

Breath <- Breath |>
  mutate(
    timepoint_use = map_chr(time_from_start,
                            assign_timepoint,
                            timepoints = timepoints_BR),
    timepoint_use = fct_relevel(timepoint_use, timepoints_BR$timepoint)
  )
```

## Analysis

```{r}

```

### Exploratory Data Analysis

```{r}

```

### Data Analysis

```{r}

```

### Results

### Discussion

## Conclusion
