---
title: "CS01: Biomarkers of Recent Use"
author: "Derek, Eddie, Rainer"
output: 
  html_document:
    toc: true
    toc_float: true
---

## Introduction

```{r setup, include=FALSE}

# control global Rmd chunk settings
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)

source("src/functions.R")
source("src/sen&spe.R")
source("src/ROC.R")
source("src/cutoff.R")

```

### Load packages

```{r load-packages, message=FALSE}

library(tidymodels)
library(tidyverse)
library(dplyr)
library(ggplot2)
library(janitor)
library(purrr)
library(rstatix)
library(cowplot)

```

## Question
-   Which compound, in which matrix, and at what cutoff is the best biomarker of recent use? (recent use is defined as 3h)
-   "do some compound repsond more to high dose/low dose compared to the other?"

## The Data

### Data Import

```{r importing-data}

WB = read.csv("data/Blood.csv")
BR = read.csv("data/Breath.csv")
OF = read.csv("data/OF.csv")


```

### Data Wrangling

We re-coded and re-leveled variables (`Treatment` and `Group`), cleans column names, and renames specific columns (`x11_oh_thc` to `thcoh`, `thc_v` to `thcv`, `thccooh_gluc` to `thc_cooh_gluc`, and `thccooh` to `thc_cooh`) in all 3 tables. Using `janitor` package to organized column names.

**Suggestion: add a new column recording whether someone should be determined to have recent THC use (THC group, within 3 hr of smoking) or not(all placebo group, or THC group outside the 3hr window)

```{r re-leveling-data}

OF <- OF |>
  mutate(Treatment = fct_recode(Treatment, 
                                "5.9% THC (low dose)" = "5.90%",
                                "13.4% THC (high dose)" = "13.40%"),
         Treatment = fct_relevel(Treatment, "Placebo", "5.9% THC (low dose)"),
         Group = fct_recode(Group, 
                            "Occasional user" = "Not experienced user",
                            "Frequent user" = "Experienced user" )) |>  
  janitor::clean_names() |>
  rename(thcoh = x11_oh_thc,
         thcv = thc_v)

WB <- WB |> 
  mutate(Treatment = fct_recode(Treatment, 
                                "5.9% THC (low dose)" = "5.90%",
                                "13.4% THC (high dose)" = "13.40%"),
         Treatment = fct_relevel(Treatment, "Placebo", "5.9% THC (low dose)")) |> 
  janitor::clean_names() |>
  rename(fluid = fluid_type,
         thcoh = x11_oh_thc,
         thccooh = thc_cooh,
         thccooh_gluc = thc_cooh_gluc,
         thcv = thc_v)

BR <- BR |> 
  mutate(Treatment = fct_recode(Treatment, 
                                "5.9% THC (low dose)" = "5.90%",
                                "13.4% THC (high dose)" = "13.40%"),
         Treatment = fct_relevel(Treatment, "Placebo", "5.9% THC (low dose)"),
         Group = fct_recode(Group, 
                            "Occasional user" = "Not experienced user",
                            "Frequent user" = "Experienced user" )) |> 
  janitor::clean_names() |> 
  rename(thc = thc_pg_pad)


compounds_WB <-  as.list(colnames(Filter(function(x) !all(is.na(x)), WB[6:13])))
compounds_BR <-  as.list(colnames(Filter(function(x) !all(is.na(x)), BR[6])))
compounds_OF <-  as.list(colnames(Filter(function(x) !all(is.na(x)), OF[6:12])))
```

Created 3 tables based on specific minutes and labeled accordingly, covering pre-smoking and subsequent post-smoking time periods for blood, breath, and oral fluid data.

```{r table-creation}

timepoints_WB <- tibble(
  start = c(-400, 0, 30, 70, 100, 180, 210, 240, 270, 300),
  stop = c(
    0,
    30,
    70,
    100,
    180,
    210,
    240,
    270,
    300,
    max(WB$time_from_start, na.rm = TRUE)
  ),
  timepoint = c(
    "pre-smoking",
    "0-30 min",
    "31-70 min",
    "71-100 min",
    "101-180 min",
    "181-210 min",
    "211-240 min",
    "241-270 min",
    "271-300 min",
    "301+ min"
  )
)

timepoints_BR <- tibble(
  start = c(-400, 0, 40, 90, 180, 210, 240, 270),
  stop = c(
    0,
    40,
    90,
    180,
    210,
    240,
    270,
    max(BR$time_from_start, na.rm = TRUE)
  ),
  timepoint = c(
    "pre-smoking",
    "0-40 min",
    "41-90 min",
    "91-180 min",
    "181-210 min",
    "211-240 min",
    "241-270 min",
    "271+ min"
  )
)

timepoints_OF <- tibble(
  start = c(-400, 0, 30, 90, 180, 210, 240, 270),
  stop = c(0, 30, 90, 180, 210, 240, 270,
           max(OF$time_from_start, na.rm = TRUE)),
  timepoint = c(
    "pre-smoking",
    "0-30 min",
    "31-90 min",
    "91-180 min",
    "181-210 min",
    "211-240 min",
    "241-270 min",
    "271+ min"
  )
)

assign_timepoint <- function(x, timepoints) {
  if (!is.na(x)) {
    timepoints$timepoint[x > timepoints$start & x <= timepoints$stop]
  } else{
    NA
  }
}
```

We created a new column, **`timepoint_use`**, in each table by mapping the **`time_from_start`** values to specific timepoints defined in separate reference data frames (**`timepoints_WB`**, **`timepoints_OF`**, **`timepoints_BR`**). Finally, releveled the **`timepoint_use`** factor variable to align with the order specified in the reference data frames. This ensures consistent and meaningful timepoint labels for subsequent analyses or visualizations in the study.

```{r}

 WB <- WB |> 
  mutate(timepoint_use = map_chr(time_from_start, 
                                 assign_timepoint, 
                                 timepoints=timepoints_WB),
         timepoint_use = fct_relevel(timepoint_use, timepoints_WB$timepoint))

OF <- OF |> 
  mutate(timepoint_use = map_chr(time_from_start, 
                                 assign_timepoint, 
                                 timepoints=timepoints_OF),
         timepoint_use = fct_relevel(timepoint_use, timepoints_OF$timepoint))

BR <- BR |> 
  mutate(timepoint_use = map_chr(time_from_start, 
                                 assign_timepoint, 
                                 timepoints=timepoints_BR),
         timepoint_use = fct_relevel(timepoint_use, timepoints_BR$timepoint))

```

remove duplicate id

```{r}

WB <- drop_dups(WB)
OF <- drop_dups(OF)
BR <- drop_dups(BR)

```

### Exploratory Data Analysis
## compounds measurements over time by treatment
heres all compounds plotted against time, coloured by group
```{r}
scatter_WB <- map(compounds_WB, ~ compound_scatterplot_group_by_treatment( 
    dataset=WB, 
    compound=.x, 
    timepoints=timepoints_WB))

scatter_OF <- map(compounds_OF, ~ compound_scatterplot_group_by_treatment( 
    dataset=OF, 
    compound=.x, 
    timepoints=timepoints_OF))

scatter_BR <- map(compounds_BR, ~ compound_scatterplot_group_by_treatment( 
    dataset=BR, 
    compound=.x, 
    timepoints=timepoints_BR))
```

heres all compounds plotted against time, coloured by treatment, log transformed
```{r, cache = TRUE}
scatter_WB_by_treatment <- map(compounds_WB, ~ compound_scatterplot_group_by_treatment_log( 
    dataset=WB, 
    compound=.x, 
    timepoints=timepoints_WB))

scatter_OF_by_treatment <- map(compounds_OF, ~ compound_scatterplot_group_by_treatment_log( 
    dataset=OF, 
    compound=.x, 
    timepoints=timepoints_OF))

scatter_BR_by_treatment <- map(compounds_BR, ~ compound_scatterplot_group_by_treatment_log( 
    dataset=BR, 
    compound=.x, 
    timepoints=timepoints_BR))
```

calculating sensitivity and specificity
```{r, cache=TRUE}
output_WB <- map_dfr(compounds_WB, 
                     ~sens_spec_cpd(dataset = WB, cpd = all_of(.x), 
                                    timepoints = timepoints_WB)) |> clean_gluc()

output_BR <- map_dfr(compounds_BR, 
                     ~sens_spec_cpd(dataset = BR,  cpd = all_of(.x),
                                    timepoints = timepoints_BR))  %>% clean_gluc()

output_OF <- map_dfr(compounds_OF, 
                     ~sens_spec_cpd(dataset = OF, cpd = all_of(.x),
                                    timepoints = timepoints_OF))  %>% clean_gluc()
```
## cutoff vs. sensitivity/specificity
Here we plot the value of the cutoff against sensitivity and specificity for every compound in every matrix, and arrange them all into one big plot
```{r, cache=TRUE}
#plot detection limit(cutoff) vs. sensitivity and specificity
ss_WB <- ss_plot(output_WB, tpts = length(unique(output_WB$time_start)), tissue = "Blood")
ss_OF <- ss_plot(output_OF, tpts = length(unique(output_OF$time_start)), tissue = "Oral Fluid")
ss_BR <- ss_plot(output_BR, tpts = length(unique(output_BR$time_start)), tissue = "Breath")


#arranges ss plots into one
ss_bottom_row <- plot_grid(ss_OF, ss_BR, labels = c('B', 'C'), label_size = 12, ncol = 2, rel_widths = c(0.66, .33))
plot_grid(ss_WB, ss_bottom_row, labels = c('A', ''), label_size = 12, ncol = 1)

```
## sensitivity vs. specificity
Here we plot sensitivity vs. specificity for every compound in every matrix, and arrange them all into one big plot
```{r}
#plot sensitivity vs. specificity
roc_WB = roc_plot(output_WB, tpts = length(unique(output_WB$time_start)), tissue = "Blood")
roc_OF = roc_plot(output_WB, tpts = length(unique(output_WB$time_start)), tissue = "Oral Fluid")
roc_BR = roc_plot(output_WB, tpts = length(unique(output_WB$time_start)), tissue = "Breath")

#arrange roc plots
roc_bottom_row <- plot_grid(roc_OF, roc_BR, labels = c('B', 'C'), label_size = 12, ncol = 2, rel_widths = c(0.66, .33))
plot_grid(roc_WB, roc_bottom_row, labels = c('A', ''), label_size = 12, ncol = 1)
```

## plot sensitivity and speciticity over time given specific cutoffs
```{r}
#pass specific cutoff into splits parameter
WB_THC <- sens_spec_cpd(dataset = WB, cpd = 'thc',
                        timepoints = timepoints_WB,
                        splits =  c(0.5, 1, 2, 5, 10)) %>% clean_gluc()

WB_THC
```
```{r}
blood_levels <- c("pre-smoking\nN=189", "0-30\nmin\nN=187", "31-70\nmin\nN=165",
                  "71-100\nmin\nN=157", "101-180\nmin\nN=168", "181-210\nmin\nN=103",
                  "211-240\nmin\nN=127", "241-270\nmin\nN=137", "271-300\nmin\nN=120",
                  "301+\nmin\nN=88")
of_levels <- c("pre-smoking\nN=192", "0-30\nmin\nN=192", "31-90\nmin\nN=117",
               "91-180\nmin\nN=99", "181-210\nmin\nN=102", "211-240\nmin\nN=83",
               "241-270\nmin\nN=90",  "271+\nmin\nN=76")


plot_cutoffs(dataset=WB_THC, 
             timepoint_use_variable=WB$timepoint_use, 
             tissue="Blood", 
             cpd="THC", 
             x_labels=blood_levels)

```


## extended question  
### wrangling  
```{r}
WB_long = WB |>
  pivot_longer(6:13, names_to = "compound")

OF_long = OF |>
  pivot_longer(6:12, names_to = "compound")
  
BR_long <- BR |> pivot_longer(6)

df_full <- bind_rows(WB_long, OF_long, BR_long)
```

### Part 1: so uh the "placebo groups might have different compound measurements" hypothesis has been disproven :(

This one is me testing out an extended question our group proposed: does taking a placebo have an effect on any of the compound? 

Theoretically, no, it shouldn't. But the graph about subjective highness being basically random really peaked my interest and I thought there might be something worth looking into. But the data disproves that theory :(

```{r}
WB_long|>
  filter(treatment == "Placebo") |>
  filter(timepoint_use == "pre-smoking" | timepoint_use == "0-30 min") |>
  ggplot(mapping = aes(x = log(value))) +
    geom_histogram(binwidth = 0.5) +
    facet_grid(vars(timepoint_use), vars(compound), scales = "free")


```
###part 2: BUT i have a different extended questgion we can look into

This one is SUPER interesting. Some of the pairplots we went over in lecture notes 12 looked like they has two separate lines, so I thought one of the variables might change how compounds correlate with each other - and they do! Namely, the low dose and high dose group seems to have different slopes when it comes to the correlation between certain compounds. 

Here's a few of the more obvious ones:
```{r}
ggplot(data = WB, 
       aes(y = cbg, x = thc, color = treatment)) +
  geom_point(alpha = 0.5) +
  facet_wrap(~group) 

ggplot(data = OF, 
       aes(y = thc, x = cbg, color = treatment)) +
  geom_point(alpha = 0.5) +
  facet_wrap(~group) 

ggplot(data = OF, 
       aes(y = cbn, x = cbg, color = treatment)) +
  geom_point(alpha = 0.5) +
  facet_wrap(~group) 

ggplot(data = OF, 
       aes(y = cbg, x = thcv, color = treatment)) +
  geom_point(alpha = 0.5) +
  facet_wrap(~group) 


```

this seems to show that even though low dose/high dose doesn't effect chemical concentration overall (as shown in the plots in lecture notes), they do change the correlation between chemicals. Maybe a potention extended question would be "do some compound repsond more to high dose/low dose compared to the other?"
```

## Analysis

```{r}


```

### Data Analysis

```{r}

```

### Results

### Discussion

## Conclusion